---
title: "Introduction to tidyverse - solutions to exercises"
date: "January 2020"
author: "Matej Piculin"
output:
  prettydoc::html_pretty:
  theme: architect
toc: yes
highlight: github
---
  
  # Solutions
  
  ## Exercise 1
  
  A dataset `who` in R has turbecolusis data from World Health Organization. You can get a short description of the data by writing `?who` in console.

The data is in wide format. Transform it into long format so that you obtain a table with attributes: country, iso2, iso3, year, group, cases.

### Solution 1

Solution by providing attribute names to gather.

```{r, eval=TRUE}
who %>%
  gather(key = group, value = cases, new_sp_m014:newrel_f65)
```

Solution by provinding names of attributes to ignore.

```{r, eval=TRUE}
who %>%
  gather(group, cases, -c(country, year, iso2, iso3))
```

Removing NA cases to shorten the table.

```{r, eval=TRUE}
who %>%
  gather(group, cases, -c(country, year, iso2, iso3)) %>%
  drop_na()
```

Or raplacing NA values with 0.

```{r, eval=TRUE}
who %>%
  gather(group, cases, -c(country, year, iso2, iso3)) %>%
  replace_na(list(cases = 0))
```

## Exercise 2

File cars.csv includes a dataset from UCI Machine Learning repository (http://mlr.cs.umass.edu/ml/datasets/Automobile).

a) Read the data into a tibble. Take care of missing values.
b) Make a linear model that predicts the price based on all numeric attributes.
c) Create a pipeline that constructs a linear model for price for each type of body_style.

### Solution 2

a) Read the data into a tibble. Take care of missing values.

```{r, eval=TRUE}
cars <- read_csv(".\\data\\cars.csv", na = "?")
```

b) Make a linear model that predicts the price based on all numeric attributes.


A solution where we extract the names of numeric attributes 'by hand' and feed them to select function.

```{r, eval=TRUE}
run_lm <- function(x){lm(price ~ ., x)}
cars %>% 
  select(names(.)[map_lgl(., is.numeric)]) %>% 
  run_lm()
```

A simple solution with `select_if()` used for such cases.

```{r, eval=TRUE}
run_lm <- function(x){lm(price ~ ., x)}
cars %>% 
  select_if(is.numeric) %>% 
  run_lm()
```

c) Create a pipeline that constructs a linear model for price for each type of body-style.

```{r, eval=TRUE}
run_lm <- function(x){lm(price ~ ., x)}
cars %>% 
  select("body-style", names(.)[map_lgl(., is.numeric)]) %>% 
  group_by(`body-style`) %>%
  nest() %>%
  mutate(lm_models = map(data, run_lm)) %>%
  .$lm_models
```

Or without complex `select()`.

```{r, eval=TRUE}
run_lm <- function(x){x %>% select_if(is.numeric) %>% lm(price ~ ., .)}
cars %>% 
  group_by(`body-style`) %>%
  nest() %>%
  mutate(lm_models = map(data, run_lm)) %>%
  .$lm_models
```